---
title: "chapter 5 Summarization Lab B"
author: "Odlin Montila "
date: ""
output: html_document
---

##### Remember to change the `author: ` field on this Rmd file to your own name.

We'll begin by doing all the same data processing as in lecture.
```{r}
library(MASS)

# Assign more descriptive variable names
colnames(birthwt) <- c("birthwt.below.2500", "mother.age", "mother.weight", 
    "race", "mother.smokes", "previous.prem.labor", "hypertension", "uterine.irr", 
    "physician.visits", "birthwt.grams")

# Assign more descriptive factor levels and convert variables to factors as needed
library(plyr)
birthwt <- transform(birthwt, 
            race = as.factor(mapvalues(race, c(1, 2, 3), 
                              c("white","black", "other"))),
            mother.smokes = as.factor(mapvalues(mother.smokes, 
                              c(0,1), c("no", "yes"))),
            hypertension = as.factor(mapvalues(hypertension, 
                              c(0,1), c("no", "yes"))),
            uterine.irr = as.factor(mapvalues(uterine.irr, 
                              c(0,1), c("no", "yes"))),
            birthwt.below.2500 = as.factor(mapvalues(birthwt.below.2500,
                              c(0,1), c("no", "yes")))
            )
```

#### 1. aggregate() vs tapply()

One of the advantages of aggregate() is that it makes it easier to view summary tables when grouping on more than two factors.  

**(a)** Use the `tapply()` function to calculate mean `birthwt.grams` grouped by race, mother's smoking status, and hypertension.

```{r}
with(birthwt, tapply(birthwt.grams, list(race, mother.smokes, hypertension), FUN = mean))
```

**(b)** Repeat part (a), this time using the `aggregate()` function.

```{r}
aggregate(birthwt.grams ~ race + mother.smokes + hypertension, FUN = mean, data = birthwt)
```



#### 2. by() practice

In today's lecture we used the `by()` function to calculate the correlation between mother's age and birth weight broken down by mother's smoking status.  Here's the command that was used:

```{r}
by(data = birthwt[c("birthwt.grams", "mother.age")], 
   INDICES = birthwt["mother.smokes"], 
   FUN = function(x) {cor(x[,1], x[,2])})
```

**(a)** Use the `by()` function to calculate the correlation between mother's age and birth weight broken down by race.

```{r}
by(data = birthwt[c("birthwt.grams", "mother.age")], 
   INDICES = birthwt["race"], 
   FUN = function(x) {cor(x[,1], x[,2])})
```

**(b)** Does the correlation appear to vary by race?



#### 3. Scatterplot practice

In class we made a pretty nice scatterplot using the following command.

```{r, fig.align='center', eval=FALSE}
# Make sure that the color-blind palette is defined
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

points.base <- c(16, 17) # 16: circle, 17: triangle
cols.base <- cbPalette[2:3] # blue and orange points
point.symbols <- points.base[birthwt$mother.smokes] 
point.cols <- cols.base[birthwt$mother.smokes]
with(birthwt, plot(mother.age, birthwt.grams, 
                   main="Birth Weight by Mother's Age", 
                   xlab="Mother's Age (Years)", 
                   ylab="Birth Weight (Grams)",
                   col=point.cols,
                   pch=point.symbols))
legend("bottomright", c("Nonsmoker","Smoker"), col=cols.base, pch=points.base)
```

**(a)** Try to create a similar plot, but instead of using different colors and symbols based on the value of `mother.smokes`, do so based on the value of `race`.  

```{r, fig.height=6, fig.width=6, fig.align='center'}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

points.base <- c(16, 17, 18)
cols.base <- cbPalette[1:3] 
point.symbols <- points.base[birthwt$race]  
point.cols <- cols.base[birthwt$race]       
with(birthwt, plot(mother.age, birthwt.grams, 
                   main="Birth Weight by Mother's Age", 
                   xlab="Mother's Age (Years)", 
                   ylab="Birth Weight (Grams)",
                   col=point.cols,
                   pch=point.symbols))
legend("bottomright", legend = levels(birthwt$race), col=cols.base,pch=points.base)
abline(h = 2500, col = cbPalette[1], lty = 3,lwd = 2)
```

**(b)** Use the `abline` command to add a horizontal line to the plot at `birthwt.grams = 2500`. 

```{r, fig.height=6, fig.width=6, fig.align='center'}
#the solution is in part a above
```