---
title: Chapter 1 -  Introduction and Basics
author: M Affouf
date: January 2, 2018
font-family: Gill Sans
output:
  beamer_presentation: default
  ioslides_presentation:
    css: ../../styles.css    
---



##Agenda
 
 
  
- Course overview 

- Introduction to R, RStudio and R Markdown

- Programming basics


##How this class will work
- No programming knowledge presumed

- Some prior stat assigned reading presumed.  
  - Basic stat definitions   
  - Hypothesis testing (t-tests, confidence intervals)
  - Linear regression

- Class attendance is mandatory

- Class will be _very_ cumulative


##Mechanics
- One session a week: 
  - First half: loading materials, concepts, methods, examples
  - Second half: Labs
- Class participation 
- Weekly Assignment
- Final project (Last 3 weeks) 
    

##Mechanics
- __Attendance and Class participation__ (20%) 
- **Labs and assignments**: Each lecture has an __accompanying lab assignment__ (40%).
 
  - Late Assignments/homework __will not be accepted for credit__

- __Final project__ (40%)
  - You will write a report analysing a policy question using a publicly available data set

##Course resources

- Google Drive will be used to post
all course notes, labs, assignments, data and announcement
- The instructor GoogleDrive: maths3141@gmail.com, I will share a folder with your email
- Every student should create a folder in their Google Drive named:
FDA_YourLastName 
and share it with maths3141@gmail.com
- All graded Labs/Assignemts should be uploaded to your FDA folder


 _highly recommended_:
  - Garrett Grolemund and Hadley Wickham, _R for Data Science_
  
  - Phil Spector, _Data Manipulation with R_
  
  - Paul Teetor, _The R Cookbook_
  
  - Winston Chang, _The R Graphics Cookbook_
  
  - Norman Matloff, _The Art of R Programming: A Tour of Statistical Software Design_
  


##Goal of this class
> This class will teach you to use R to:    
- Generate graphical and tabular data summaries    
- Perform statistical analyses (e.g., hypothesis testing, regression modeling)   
- Handle different data sources: Surveys, Financial, and text data   
- Produce _reproducible_ statistical reports using R Markdown
   


##Why R?
- Free (open-source)
- Programming language (not point-and-click)
- Excellent graphics
- Offers broadest range of statistical tools
- Easy to generate reproducible reports
- Easy to integrate with other tools



##The R Console

Basic interaction with R is through typing in the **console**



##The R Console
- You type in commands, R gives back answers (or errors)

- Menus and other graphical interfaces are extras built on top of the console

- We will use **RStudio** in this class

1. Download R: 

2. Then download RStudio: http://www.rstudio.com/


##


**RStudio** is an IDE for R

RStudio has 4 main windows ('panes'):
- Source
- Console 
- Workspace/History
- Files/Plots/Packages/Help


- Use the **Console** pane to type or paste commands to get output from R

- To look up the help file for a function or data set, type `?function` into the Console
  - E.g.,  try typing in `?mean`
  
- Use the `tab` key to auto-complete function and object names


- Use the **Source** pane to create and edit R and Rmd files
- The menu bar of this pane contains handy shortcuts for sending code to the **Console** for evaluation


- By default, any figures you produce in R will be displayed in the **Plots** tab 
  - Menu bar allows you to Zoom, Export, and Navigate back to older plots
- When you request a help file (e.g., `?mean`), the documentation will appear in the **Help** tab


##RStudio: Panes overview

1. __Source__ pane: create a file that you can save and run later

2. __Console__ pane: type or paste in commands to get output from R
3. __Workspace/History__ pane: see a list of variables or previous commands
4.  __Files/Plots/Packages/Help__ pane: see plots, help pages, and other items in this window.


## Getting Started

* You should have the latest version of R installed!
* Open R Studio
* Files --> New --> R Script
* Save the blank R script as "day1.R" in a directory of your choosing
* Add a comment header

## Commenting in Scripts

Add a comment header to day1.R :`#` is the comment symbol

```{r}
#################
# Title: Demo R Script
# Author: M Affouf
# Date: 12/24/2017
# Purpose: Demonstrate comments in R
###################
 
# nothing to its right is evaluated

# this # is still a comment
### you can use many #'s as you want

# sometimes you have a really long comment,
#    like explaining what you are doing 
#    for a step in analysis. 
# Take it to another line
```

## Explaining output on slides

In slides, a command (we'll also call them code or a code chunk) will look like this
```{r code}
print("I'm code")
```

And then directly after it, will be the output of the code.  
So `print("I'm code")` is the code chunk and [1] "I'm code" is the output.

## R as a calculator

```{r calcDemo}
2 + 2
2 * 4
2 ^ 3
```

Note, when you type your command, R inherently thinks you want to print the result. 

## R as a calculator

* The R console is a full calculator
* Try to play around with it:
    * +, -, /, * are add, subtract, divide and multiply
    * ^ or ** is power
    * parentheses -- ( and ) -- work with order of operations 

## R as a calculator

```{r calcDemo2}
2 + (2 * 3)^2
(1 + 3) / 2 + 45
```

## R as a calculator

Try evaluating the following:

* `2 + 2 * 3 / 4 -3`
* `2 * 3 / 4 * 2`
* `2^4 - 1`

---

* You can create variables from within the R environment and from files on your computer
* R uses "=" or "<-" to assign values to a variable name
* Variable names are case-sensitive, i.e. X and x are different

```{r assign}
x = 2 # Same as: x <- 2
x
x * 4
x + 2
```


## R variables

* The most comfortable and familiar class/data type for many of you will be `data.frame`
* You can think of these as essentially Excel spreadsheets with rows (usually subjects or observations) and columns (usually variables)

* Go to RStudio --> Tools --> Import Dataset --> From Web URL then paste

http://data/Charm_City_Circulator_Ridership.csv

```{r, echo = FALSE}
Charm_City_Circulator_Ridership <- read.csv("Charm_City_Circulator_Ridership.csv")
```

## R variables {.smaller}
* We can display the top of the data with `head`:

```{r df1,  comment=""}
head(Charm_City_Circulator_Ridership)
```


## R variables
* `data.frame`s are somewhat advanced objects in R; we will start with simpler objects;
* Here we introduce "1 dimensional" classes; these are often referred to as 'vectors'
* Vectors can have multiple sets of observations, but each observation has to be the same class.

```{r assignClass}
class(x)
y = "hello world!"
print(y)
class(y)
```

## R variables

Try assigning your full name to an R variable called `name`

## R variables

Try assigning your full name to an R variable called `name`

```{r myName}
name = "Rober De Niro"
name
```

## The 'combine' function

The function `c()` collects/combines/joins single R objects into a vector of R objects. It is mostly used for creating vectors of numbers, character strings, and other data types. 

```{r assign3a}
x <- c(1, 4, 6, 8)
x
class(x)
```

## The 'combine' function

Try assigning your first and last name as 2 separate character strings into a single vector called `name2`

## The 'combine' function

Try assigning your first and last name as 2 separate character strings into a length-2 vector called `name2`

```{r myName2}
name2 = c("Robert","De Niro")
name2
```

## R variables

`length()`: Get or set the length of vectors (including lists) and factors, and of any other R object for which a method has been defined.

```{r assign3b}
length(x)
y
length(y)
```

##  R variables

What do you expect for the length of the `name` variable? What about the `name2` variable?

What are the lengths of each?

##  R variables

What do you expect for the length of the `name` variable? What about the `name2` variable?

What are the lengths of each?

```{r myName3}
length(name)
length(name2)
```

## R variables

You can perform functions to entire vectors of numbers very easily.

```{r assign4}
x + 2
x * 3
x + c(1, 2, 3, 4)
```

## R variables

But things like algebra can only be performed on numbers.

```
> name2 + 4
[1] Error in name2 * 4 : non-numeric argument
 to binary operator
```

## R variables

And save these modified vectors as a new vector.

```{r assign5}
y = x + c(1, 2, 3, 4)
y 
```

Note that the R object `y` is no longer "Hello  World!" - It has effectively been overwritten by assigning new data to the variable

## R variables

* You can get more attributes than just class.  The function `str` gives you the structure of the object.

```{r assign2}
str(x)
str(y)
```
This tells you that `x` is a numeric vector and tells you the length.

## Back to our `data.frame` example {.smaller}

* Let's see what the structure of our `data.frame` is:

```{r str_charm}
str(Charm_City_Circulator_Ridership)
```

## Review
* Creating a new script
* Using R as a calculator
* Assigning values to variables
* Performing algebra on numeric variables





##R Markdown
- R Markdown allows the user to integrate R code into a report

- When data changes or code changes, so does the report

- No more need to copy-and-paste graphics, tables, or numbers

- Creates __reproducible__ reports
  - Anyone who has your R Markdown (.Rmd) file and input data can re-run your analysis and get the exact same results (tables, figures, summaries)
  
- Can output report in HTML (default), Microsoft Word, or PDF

- This example shows an **R Markdown** (.Rmd) file opened in the Source pane of RStudio.  
- To turn an Rmd file into a report, click the **Knit HTML** button in the Source pane menu bar
- The results will appear in a **Preview window**, as shown on the right
- You can knit into html (default), MS Word, and pdf format
- These lecture slides are also created in RStudio (R Presentation)


- To integrate R output into your report, you need to use R code chunks
- All of the code that appears in between the "triple back-ticks" gets executed when you Knit

##In-class exercise: Hello world!

1. Open **RStudio** on your machine

2. File > New File > R Markdown ...
3. Change `summary(cars)` in the first code block to `print("Hello world!")`
4. Click `Knit HTML` to produce an HTML file.
5. Save your Rmd file as `helloworld.Rmd`



##Basics: the class in a nutshell
- Everything we'll do comes down to applying **functions** to **data**

- **Data**:  things like 7, "seven", $7.000$, the matrix $\left[ \begin{array}{ccc} 7 & 7 & 7 \\ 7 & 7 & 7\end{array}\right]$

- **Functions**: things like $\log{}$, $+$ (two arguments), $<$ (two), $\mod{}$ (two), `mean` (one)

> A function is a machine which turns input objects (**arguments**) into an output object (**return value**), possibly with **side effects**, according to a definite rule

##Data building blocks

You'll encounter different kinds of data types

- **Booleans** Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero)
- **Characters** fixed-length blocks of bits, with special coding;
**strings** = sequences of characters
- **Floating point numbers**: a fraction (with a finite number of bits) times an exponent, like $1.87 \times {10}^{6}$
- **Missing or ill-defined values**: `NA`, `NaN`, etc.


##Operators (functions)
You can use R as a very, very fancy calculator

Command | Description
--------|-------------
`+,-,*,\` | add, subtract, multiply, divide
`^` | raise to the power of
`%%` | remainder after division (ex: `8 %% 3 = 2`)
`( )` | change the order of operations
`log(), exp()` | logarithms and exponents (ex: `log(10) = 2.302`)
`sqrt()` | square root
`round()` | round to the nearest whole number (ex: `round(2.3) = 2`)
`floor(), ceiling()` | round down or round up 
`abs()` | absolute value
###

```{r}
7 + 5 # Addition
7 - 5 # Subtraction
7 * 5 # Multiplication
7 ^ 5 # Exponentiation
```
###
```{r}
7 / 5 # Division
7 %% 5 # Modulus
7 %/% 5 # Integer division 
```

##Operators cont'd.
  **Comparisons** are also binary operators; they take two objects, like numbers, and give a Boolean
```{r}
7 > 5
7 < 5
7 >= 7
7 <= 5
```
###
  ```{r}
7 == 5
7 != 5
```

##Boolean operators
  Basically "and" and "or":
  
```{r}
(5 > 7) & (6*7 == 42)
(5 > 7) | (6*7 == 42)
```

<small>(will see special doubled forms, `&&` and `||`, later)</small>
  
  
##  More types
  
- `typeof()` function returns the type

- `is.`_foo_`()` functions return Booleans for whether the argument is of type _foo_

- `as.`_foo_`()` (tries to) "cast" its argument to type _foo_ --- to translate it sensibly into a _foo_-type value

<small>**Special case**: `as.factor()` will be important later for telling R when numbers are actually encodings and not numeric values.  (E.g., 1 = High school grad; 2 = College grad; 3 = Postgrad) </small>
###
  ```{r}
typeof(7)
is.numeric(7)
is.na(7)
```
###  
  ```{r}
is.character(7)
is.character("7")
is.character("seven")
is.na("seven")
```
  
##Variables
  
  We can give names to data objects; these give us **variables**
  
  A few variables are built in:
```{r}
pi
```

Variables can be arguments to functions or operators, just like constants:
```{r}
pi*10
cos(pi)
```

##Assignment operator
Most variables are created with the **assignment operator**, `<-` or `=`  

```{r}
time.factor <- 12
time.factor
time.in.years = 2.5
time.in.years * time.factor
```
###
The assignment operator also changes values:
```{r}
time.in.months <- time.in.years * time.factor
time.in.months
time.in.months <- 45
time.in.months
```
###
  
- Using names and variables makes code: easier to design, easier to debug, less prone to bugs, easier to improve,  and easier for others to read

- Avoid "magic constants"; use named variables

- Use descriptive variable names
  - Good: `num.students <- 35`
  - Bad: `ns <- 35 `

  
  
##  The workspace
  What names have you defined values for?
```{r}
ls()
```

Getting rid of variables:
```{r}
rm("time.in.months")
ls()
```

##First data structure: vectors
  
- Group related data values into one object, a **data structure**
  
- A **vector** is a sequence of values, all of the same type

- `c()` function returns a vector containing all its arguments in order
```{r}
students <- c("Sean", "Louisa", "Frank", "Farhad", "Li")
midterm <- c(80, 90, 93, 82, 95)
```

- Typing the variable name at the prompt causes it to display

```{r}
students
```

##Indexing
- `vec[1]` is the first element, `vec[4]` is the 4th element of `vec`
```{r}
students
students[4]
```
- `vec[-4]` is a vector containing all but the fourth element
```{r}
students[-4]
```


##Vector arithmetic
  
Operators apply to vectors "pairwise" or "elementwise":
```{r}
final <- c(78, 84, 95, 82, 91) # Final exam scores
midterm # Midterm exam scores
midterm + final # Sum of midterm and final scores
(midterm + final)/2 # Average exam score
course.grades <- 0.4*midterm + 0.6*final # Final course grade
course.grades
```

##Pairwise comparisons
Is the final score higher than the midterm score?
```{r}
midterm 
final
final > midterm
```

Boolean operators can be applied elementwise:
```{r}
(final < midterm) & (midterm > 80)
```

##Functions on vectors
Command | Description
--------|------------
`sum(vec)` | sums up all the elements of `vec`
`mean(vec)` | mean of `vec`
`median(vec)` | median of `vec`
`min(vec), max(vec)` | the largest or smallest element of `vec`
`sd(vec), var(vec)` | the standard deviation and variance of `vec`
`length(vec)` | the number of elements in `vec`
`pmax(vec1, vec2), pmin(vec1, vec2)` | example: `pmax(quiz1, quiz2)` returns the higher of quiz 1 and quiz 2 for each student
`sort(vec)` | returns the `vec` in sorted order
`order(vec)` | returns the index that sorts the vector `vec`
`unique(vec)` | lists the unique elements of `vec`
`summary(vec)` | gives a five-number summary  
`any(vec), all(vec)` | useful on Boolean vectors

##Functions on vectors
```{r}
course.grades
mean(course.grades) # mean grade
median(course.grades)
sd(course.grades) # grade standard deviation
```

##More functions on vectors
```{r}
sort(course.grades)
max(course.grades) # highest course grade
min(course.grades) # lowest course grade
```
##Referencing elements of vectors
```{r}  
students
```
  
  Vector of indices:
```{r}
students[c(2,4)]
```

Vector of negative indices
```{r}
students[c(-1,-3)]
```

##More referencing

`which()` returns the `TRUE` indexes of a Boolean vector:
```{r}
course.grades
a.threshold <- 90 # A grade = 90% or higher
course.grades >= a.threshold # vector of booleans
a.students <- which(course.grades >= a.threshold) # Applying which() 
a.students
students[a.students] # Names of A students
```

##Named components
  
You can give names to elements or components of vectors
```{r}
students
names(course.grades) <- students # Assign names to the grades
names(course.grades)
course.grades[c("Sean", "Frank","Li")] # Get final grades for 3 students
```
Note the labels in what R prints; these are not actually part of the value

##Useful RStudio tips
Keystroke | Description
----------|-------------
`<tab>` | autocompletes commands and filenames, and lists arguments for functions. Highly useful!
`<up>` | cycle through previous commands in the console prompt
`<ctrl-up>` | lists history of previous commands matching an unfinished one
`<ctrl-enter>` | paste current line from source window to console. Good for trying things out ideas from a source file.
`<ESC>` | as mentioned, abort an unfinished command and get out of the + prompt

<br>


