---
title: "chapter2 Functions Lab B"
author: "Odlin Montila"
output: html_document
---

##### Remember to change the `author: ` field on this Rmd file to your own name.

For the first two problems we'll use the Cars93 data set from the MASS library.

```{r}
library(MASS)
head(Cars93)
```

#### 1. Manipulating data frames

Use the `transform()` and `log()` functions to create a new data frame called `Cars93.log` that has `MPG.highway` and `MPG.city` replaced with `log(MPG.highway)` and `log(MPG.city)`.

```{r}
t=transform(Cars93,MPG.highway=log(MPG.highway),MPG.city=log(MPG.city))
head(t)
```

#### 2. Table function

Use the `table()` function to tabulate the data by DriveTrain and Origin.  

```{r}
table( Cars93$Origin, Cars93$DriveTrain )
```


#### 3. Functions, lists, and if-else practice

**(a)** Write a function called `isPassingGrade` whose input `x` is a number, and which returns `FALSE` if `x` is lower than 50 and `TRUE` otherwise.

```{r}
isPassingGrade <- function(x){
  if (x<50)
    {
    Result <- "FALSE"
    }
     else
    {
    Result <- "TRUE"
  }
  return(Result)

  }
```

**(b)** Write a function called `sendMessage` whose input `x` is a number, and which prints `Congratulations` if `isPassingGrade(x)` is `TRUE` and prints `Oh no!` if `isPassingGrade(x)` is `FALSE`.

```{r}
sendMessage <- function(x){
   if (isPassingGrade(x)=="TRUE")
       {
     return("Congratulations")
   }
  else
  {
    return("Oh no!")
  }
}
```

**(c)** Write a function called `gradeSummary` whose input `x` is a number.  Your function will return a list with two elements, named `letter.grade` and `passed`.  The letter grade will be `"A"` if `x` is at least `90`.  The letter grade will be `"B"` if `x` is between `80` and `90`.  The letter grade will be `"F"` if `x` is lower than `"80"`.  If the student's letter grade is an A or B, `passed` should be TRUE; `passed` should be FALSE otherwise.

```{r}
gradeSummary <- function(x){

  
  if(x>=90)
  {
  letter.grade <- "A"
  passed <- "TRUE"
  }
  else if (x>=80)
  {
  letter.grade <- "B"
  passed <- "TRUE"
  }
  else
  {
  letter.grade <- "F"
  passed <- "FALSE"
  }
  
 summary.list <- list (letter.grade=letter.grade,passed=passed)
 return(summary.list)
}
```

To check if your function works, try the following cases:

`x = 91` should return 
```{r, echo = FALSE}
list(letter.grade = "A", passed = TRUE)
```

`x = 62` should return 
```{r, echo = FALSE}
list(letter.grade = "F", passed = FALSE)
```