---
title: "chapter 6 Data Cleaning & apply Lab B"
author: "Odlin Montila"
output: html_document
---

##### Remember to change the `author: ` field on this Rmd file to your own name.

#### 1. Loop practice

**(a)** Write a function called `calculateRowMeans` that uses a **for loop** to calculate the row means of a matrix `x`.

```{r}
calculateRowMeans <- function(x) {
  row.means <- numeric(nrow(x))
  for(i in 1:nrow(x)) {
    row.means[i] <- mean(x[i,])
  }
  row.means
}
```

**(b)** Try out your function on the random matrix `fake.data` defined below.

```{r}
set.seed(12345) # Set seed of random number generator
fake.data <- matrix(runif(800), nrow=25)
```

**(b)** Use the `apply()` function to calculate the row means of the matrix `fake.data`

```{r}
apply(fake.data, MARGIN=1, FUN=mean)
```

**(c)** Compare this to the output of the `rowMeans()` function to check that your calculation is correct.

```{r}
identical(calculateRowMeans(fake.data), apply(fake.data, MARGIN=1, FUN=mean))
```

#### 2. tapply() practice

Use `tapply()` on the Cars93 data set to create a table showing the average `Turn.circle` of cars, broken down by vehicle `Type` and `DriveTrain`

```{r}
library(MASS) 
with(Cars93, tapply(Turn.circle, INDEX = list(Type, DriveTrain), FUN=mean))
```

#### 3. with() practice

```{r}
library(MASS) # Load the library that Cars93 is in
```

Below is a code chunk that produces box plots of Horsepower against vehicle Type.  

```{r, fig.align='center', fig.width=7, fig.height=5}
boxplot(Cars93$Horsepower ~ Cars93$Type)
```

Use the `with()` function to produce the same plot with different syntax.

```{r, fig.align='center', fig.width=7, fig.height=5}
with(Cars93, boxplot(Horsepower ~ Type))

```