---
title: Note 14 - BASIC DATA VISUALIZATION
author: MA
date: Spring 18
output: 
  html_document:
    toc: true
    toc_depth: 5
    fig_width: 5
    fig_height: 5
---

## Barplots and Pie Charts
Barplots and pie charts are commonly used to visualize **qualitative data by category frequency**. 


### Building a Barplot
A barplot draws either vertical or horizontal bars, typically separated by white
space, to visualize frequencies according to the relevant categories. 

Barplot can also
be used to visualize other quantities, such as means or proportions, which
directly depend upon these frequencies.
As an example, let's use the `mtcars` data set  Detailing various characteristics of 32 classic performance cars in
the mid-1970s, the first five records can be viewed directly from the prompt.

```{r}
mtcars[1:5,]
```

The documentation in `?mtcars` explains the variables that have been
recorded. Of these, `cyl` provides the number of cylinders in each engine-
four, six, or eight. To find out how many cars were observed with each number of cylinders, you can use table, as shown here:
```{r}
cyl.freq <- table(mtcars$cyl)
cyl.freq
```


The result is easily displayed as a barplot, as shown here:
```{r}
barplot(cyl.freq) 
```

This plot displays the number of four-, six-, and eight-cylinder cars in the
data set but is admittedly rather uninteresting, and without annotations it's
not clear what's being summarized. Fortunately, it's easy to annotate such
plots and further split up the frequencies of each bar according to an additional
categorical variable. Consider the following code where, this time,
you're finding the counts associated with `cyl` by transmission `(am)`:

```{r}
table(mtcars$cyl[mtcars$am==0])
table(mtcars$cyl[mtcars$am==1])
```


If you aim to produce a barplot that's `stacked` (where bars are split up
vertically) or `dodged` (where bars are broken up and placed beside each
other), `barplot` requests its first argument as a suitably arranged matrix. You
could construct it from the previous two vectors using matrix, but it's easier
to just continue using `table`.

```{r}
cyl.freq.matrix <- table(mtcars$am,mtcars$cyl)
cyl.freq.matrix
```

```{r}
barplot(cyl.freq.matrix,beside=TRUE,horiz=TRUE,las=1,
main="Performance car counts\nby transmission and cylinders",
names.arg=c("V4","V6","V8"),legend.text=c("auto","manual"),
args.legend=list(x="bottomright"))
```

The help file `?barplot` explains the options here in detail. To label the
bars according to the categories of the column variable of the matrix that
was initially passed to barplot, you use a character vector of the appropriate
length passed to `names.arg`. The options `beside=TRUE` and `horiz=TRUE` select a
dodged, horizontal barplot. If both options were FALSE, a stacked, vertical
barplot would be selected. The argument `las=1` forces the labels on the vertical
axis to appear horizontally, rather than parallel to it. The final two arguments,
`legend.text` and `args.legend`, are used for the legend.


Similar plots may be produced using `ggplot2`. If you load the installed
package with `library("ggplot2")`

```{r}
library("ggplot2")

qplot(factor(mtcars$cyl),geom="bar")
```

```{r}
qplot(factor(mtcars$cyl),geom="blank",fill=factor(mtcars$am), xlab="",
ylab="",main="Performance car counts\nby transmission and cylinders") + 
  geom_bar(position="dodge") +
  scale_x_discrete(labels=c("V4","V6","V8")) +
  scale_y_continuous(breaks=seq(0,12,2)) + 
  theme_bw() + coord_flip() +
  scale_fill_grey(name="Trans.",labels=c("auto","manual"))
```


###A Quick Pie Chart

The venerable `pie chart` is an alternative option for visualizing frequency-based
quantities across levels of categorical variables, with appropriately
sized "slices" representing the relative counts of each categorical variable.
```{r}
pie(table(mtcars$cyl),labels=c("V4","V6","V8"),
col=c("white","gray","black"),main="Performance cars by cylinders")
```

###Histograms

The `barplot` is intuitively sensible for counting observations in relation to
categorical variables but is of virtually no use if the variable you're interested
in is numeric-continuous. To visualize the distribution of continuous measurements,
you can use a `histogram`-a tool that's sometimes confused with
a barplot owing to its similar appearance. A `histogram` also measures frequencies,
but in targeting a numeric-continuous variable, it's first necessary
to "bin" the observed data, meaning to define intervals and then count the
number of continuous observations that fall within each one. The size of this
interval is known as the `binwidth`.
For a simple example of a `histogram`, consider the horsepower data of
the 32 cars in `mtcars`, given in the fourth column, named `hp`.

```{r}
mtcars$hp
```

```{r}
hist(mtcars$hp)
```


The accuracy of a histogram as a representation of the shape of a distribution
of measurements depends solely upon the widths of the intervals
used to bin the data. Binwidths are controlled in hist by the breaks argument.
You can manually set these by supplying a vector, giving each breakpoint,
to breaks. This is done in the following code by halving the width of
each bin from 50 to 25 and widening the overall range somewhat, using an
evenly spaced sequence.

```{r}
hist(mtcars$hp,breaks=seq(0,400,25),col="gray",main="Horsepower",xlab="HP")

abline(v=c(mean(mtcars$hp),median(mtcars$hp)),lty=c(2,3),lwd=2)

legend("topright",legend=c("mean HP","median HP"),lty=c(2,3),lwd=2)
```

So, rather than defaulting to a data-driven algorithm such as hist,
ggplot2 encourages users to become aware of the issue and actively set their
own binwidths.

```{r}
qplot(mtcars$hp,geom="blank",main="Horsepower",xlab="HP") +
  geom_histogram(color="black",fill="white",breaks=seq(0,400,25), closed="right") +
  geom_vline(mapping=aes(xintercept=c(mean(mtcars$hp),median(mtcars$hp)), linetype=factor(c("mean","median"))),show.legend=TRUE) + 
  scale_linetype_manual(values=c(2,3)) +
  labs(linetype="")
```


###Box-and-Whisker Plots
An especially popular alternative to the histogram is the box-and-whisker plot,
or simply boxplot for short. This is merely a visual representation of the fivenumber
summary.


Let's return to the built-in quakes data frame of the 1,000 seismic events near
Fiji. For the sake of comparison, you can examine both a histogram and a
boxplot of the magnitudes of these events using default base R behavior.

```{r}
hist(quakes$mag)
boxplot(quakes$mag)

```

Like the histogram, a boxplot shows important features of the distribution,
such as global (in other words, overall) centrality, spread, and skewness.

###Side-by-Side Boxplots
One particularly pleasing aspect of these plots is the ease with which you
can compare the five-number summary distributions of different groups
with side-by-side boxplots. Again using the quakes data, define the following
corresponding factor and inspect the first five elements

```{r}
stations.fac <- cut(quakes$stations,breaks=c(0,50,100,150))
stations.fac[1:5]
```

Recall that the stations variable records how many monitoring stations
detected each event. This code has produced a factor breaking up these
observations into one of three groups-events detected by 50 stations or
fewer, between 51 and 100 stations, and between 101 and 150 stations. Thus,
you can compare the distributions of the magnitudes of the events according
to these three groups

```{r}
boxplot(quakes$mag~stations.fac,
xlab="# stations detected",ylab="Magnitude",col="gray")
```

Turning to ggplot2 functionality, qplot can produce the same type of plot
easily,
```{r}
qplot(stations.fac,quakes$mag,geom="boxplot",
xlab="# stations detected",ylab="Magnitude")
```


###Scatterplots
A scatterplot is most frequently used to identify a relationship between the
observed values of two different numeric-continuous variables, displayed as
x-y coordinate plots. The coordinate-wise nature of base R graphics lends
itself naturally to the creation of scatterplots

the famous iris data. Collected in the mid-
1930s, this data frame of 150 rows and 5 columns consists of petal and sepal
measurements for three species of perennial iris flowers-Iris setosa, Iris virginica,
and Iris versicolor (Anderson, 1935; Fisher, 1936). You can view the
first five records here:

```{r}
iris[1:5,]
```

Single Plot
You can modify a simple scatterplot to split the plotted points according to a
categorical variable, exposing potential differences between any visible relationships
with respect to the continuous variables. For example, using base
R graphics, you can examine the petal measurements according to the three
species.

you can manually build up this plot by first using type="n" to generate an
empty plotting region of the correct dimensions and subsequently adding
the points corresponding to each species, altering point character and color
as desired.

```{r}
plot(iris[,4],iris[,3],type="n",xlab="Petal Width (cm)",
ylab="Petal Length (cm)")

points(iris[iris$Species=="setosa",4],
iris[iris$Species=="setosa",3],pch=19,col="black")
points(iris[iris$Species=="virginica",4],
iris[iris$Species=="virginica",3],pch=19,col="gray")
points(iris[iris$Species=="versicolor",4],
iris[iris$Species=="versicolor",3],pch=1,col="black")
legend("topleft",legend=c("setosa","virginica","versicolor"),
col=c("black","gray","black"),pch=c(19,19,1))
```

Note that the Iris virginica
species has the largest petals, followed by Iris versicolor, and the smallest
petals belong to Iris setosa. However, this code, while functional, is fairly
cumbersome. You can generate the same image more simply by first setting
up vectors that specify the desired point character and color for each individual
observation.

###Matrix of Plots
The "single" type of planar scatterplot is really useful only when comparing
two numeric-continuous variables. When there are more continuous variables
of interest, it isn't possible to display this information satisfactorily on
a single plot. A simple and common solution is to generate a two-variable
scatterplot for each pair of variables and show them together in a structured
way; this is referred to as a scatterplot matrix. Making use of the iris_pch and
iris_col vectors created earlier, you can generate a scatterplot matrix for all
four continuous variables in iris, retaining the distinction between species.
Working with base R graphics, use the pairs function.
```{r}
iris_pch <- rep(19,nrow(iris))
iris_pch[iris$Species=="versicolor"] <- 1

iris_col <- rep("black",nrow(iris))
iris_col[iris$Species=="virginica"] <- "gray"


pairs(iris[,1:4],pch=iris_pch,col=iris_col,cex=0.75)
```

or

```{r}
library("GGally")

ggpairs(iris,mapping=aes(col=Species),axisLabels="internal")
```

ggpairs offers an impressive visual array for such a short line of code.
The output not only gives you the lower half of the scatterplot matrix produced
with pairs but also provides equivalent histograms along the bottom
and boxplots along the right. It also displays the estimates of the correlation
coefficients.


##Additinal examples of Basic graphs

In the following examples, you'll plot the outcome of a study investigating a new
treatment for rheumatoid arthritis. The data are contained in the Arthritis data
frame distributed with the vcd package.

In the Arthritis study, the variable Improved records the patient outcomes for individuals
receiving a placebo or drug:
```{r}
library(vcd)
counts <- table(Arthritis$Improved)
counts
```

```{r}
barplot(counts,
main="Simple Bar Plot",
xlab="Improvement", ylab="Frequency")

barplot(counts,
main="Horizontal Bar Plot",
xlab="Frequency", ylab="Improvement",
horiz=TRUE)
```

**Creating bar plots with factor variables**

If the categorical variable to be plotted is a factor or ordered factor, you can create a
vertical bar plot quickly with the plot() function. Because Arthritis$Improved is
a factor, the code
```{r}
plot(Arthritis$Improved, main="Simple Bar Plot",
xlab="Improved", ylab="Frequency")

plot(Arthritis$Improved, horiz=TRUE, main="Horizontal Bar Plot",
xlab="Frequency", ylab="Improved")
```


**Stacked and grouped bar plots**

If height is a matrix rather than a vector, the resulting graph will be a stacked or
grouped bar plot. If beside=FALSE (the default), then each column of the matrix produces
a bar in the plot, with the values in the column giving the heights of stacked
"sub-bars." If beside=TRUE, each column of the matrix represents a group, and the values
in each column are juxtaposed rather than stacked.
Consider the cross-tabulation of treatment type and improvement status:
```{r}
library(vcd)

counts <- table(Arthritis$Improved, Arthritis$Treatment)

counts


barplot(counts,
main="Stacked Bar Plot",
xlab="Treatment", ylab="Frequency",
col=c("red", "yellow","green"),
legend=rownames(counts))

barplot(counts,
main="Grouped Bar Plot",
xlab="Treatment", ylab="Frequency",
col=c("red", "yellow", "green"),
legend=rownames(counts), beside=TRUE)

```

**Mean bar plots**

Bar plots needn't be based on counts or frequencies. You can create bar plots that represent
means, medians, standard deviations, and so forth by using the aggregate function
and passing the results to the barplot() function.

```{r}
states <- data.frame(state.region, state.x77)
means <- aggregate(states$Illiteracy, by=list(state.region), FUN=mean)
means

means <- means[order(means$x),]
means

barplot(means$x, names.arg=means$Group.1)
title("Mean Illiteracy Rate")
```

**Spinograms**

Before finishing our discussion of bar plots, let's take a look at a specialized version
called a spinogram. In a spinogram, a stacked bar plot is rescaled so that the height of
each bar is 1 and the segment heights represent proportions. Spinograms are created
through the spine() function of the vcd package. The following code produces a simple
spinogram:
```{r}
library(vcd)
attach(Arthritis)
counts <- table(Treatment, Improved)
spine(counts, main="Spinogram Example")
detach(Arthritis)
```

**Pie charts**

```{r}
slices <- c(10, 12,4, 16, 8)
lbls <- c("US", "UK", "Australia", "Germany", "France")
pie(slices, labels = lbls,
main="Simple Pie Chart")
pct <- round(slices/sum(slices)*100)
lbls2 <- paste(lbls, " ", pct, "%", sep="")
pie(slices, labels=lbls2, col=rainbow(length(lbls2)),
main="Pie Chart with Percentages")

library(plotrix)
pie3D(slices, labels=lbls,explode=0.1,
main="3D Pie Chart ")
mytable <- table(state.region)
lbls3 <- paste(names(mytable), "\n", mytable, sep="")
pie(mytable, labels = lbls3,
main="Pie Chart from a Table\n (with sample sizes)")
```

```{r}
library(plotrix)
slices <- c(10, 12,4, 16, 8)
lbls <- c("US", "UK", "Australia", "Germany", "France")
fan.plot(slices, labels = lbls, main="Fan Plot")
```


In a fan plot, the slices are rearranged to overlap
each other, and the radii are modified so
that each slice is visible

**Histograms**

```{r}
par(mfrow=c(2,2))
hist(mtcars$mpg)
hist(mtcars$mpg,
breaks=12,
col="red",
xlab="Miles Per Gallon",
main="Colored histogram with 12 bins")
hist(mtcars$mpg,
freq=FALSE,
breaks=12,
col="red",
xlab="Miles Per Gallon",
main="Histogram, rug plot, density curve")
rug(jitter(mtcars$mpg))
lines(density(mtcars$mpg), col="blue", lwd=2)
x <- mtcars$mpg
h<-hist(x,
breaks=12,
col="red",
xlab="Miles Per Gallon",
main="Histogram with normal curve and box")
xfit<-seq(min(x), max(x), length=40)
yfit<-dnorm(xfit, mean=mean(x), sd=sd(x))
yfit <- yfit*diff(h$mids[1:2])*length(x)
lines(xfit, yfit, col="blue", lwd=2)
box()
```


**Kernel density plots**

`plot(density(x))`

```{r}
par(mfrow=c(2,1))
d <- density(mtcars$mpg)
plot(d)
d <- density(mtcars$mpg)
plot(d, main="Kernel Density of Miles Per Gallon")
polygon(d, col="red", border="blue")
rug(mtcars$mpg, col="brown")
```


```{r}
library(sm)
attach(mtcars)
cyl.f <- factor(cyl, levels= c(4,6,8),
labels = c("4 cylinder", "6 cylinder",
"8 cylinder"))
sm.density.compare(mpg, cyl, xlab="Miles Per Gallon")
#title(main="MPG Distribution by Car Cylinders")
colfill<-c(2:(1+length(levels(cyl.f))))
#legend(locator(1), levels(cyl.f), fill=colfill)
detach(mtcars)
```

**Box Plots**

```{r}
boxplot(mpg ~ cyl, data=mtcars,
main="Car Mileage Data",
xlab="Number of Cylinders",
ylab="Miles Per Gallon")

boxplot(mpg ~ cyl, data=mtcars,
notch=TRUE,
varwidth=TRUE,
col="red",
main="Car Mileage Data",
xlab="Number of Cylinders",
ylab="Miles Per Gallon")
```


```{r}
mtcars$cyl.f <- factor(mtcars$cyl,
levels=c(4,6,8),
labels=c("4","6","8"))
mtcars$am.f <- factor(mtcars$am,
levels=c(0,1),
labels=c("auto", "standard"))
boxplot(mpg ~ am.f *cyl.f,
data=mtcars,
varwidth=TRUE,
col=c("gold","darkgreen"),
main="MPG Distribution by Auto Type",
xlab="Auto Type", ylab="Miles Per Gallon")
```

**Violin plots**

```{r}
library(vioplot)
x1 <- mtcars$mpg[mtcars$cyl==4]
x2 <- mtcars$mpg[mtcars$cyl==6]
x3 <- mtcars$mpg[mtcars$cyl==8]
vioplot(x1, x2, x3,
names=c("4 cyl", "6 cyl", "8 cyl"),
col="gold")
```

**Dot Plots**

```{r}
dotchart(mtcars$mpg, labels=row.names(mtcars), cex=.7,
main="Gas Mileage for Car Models",
xlab="Miles Per Gallon")
```

```{r}
x <- mtcars[order(mtcars$mpg),]
x$cyl <- factor(x$cyl)
x$color[x$cyl==4] <- "red"
x$color[x$cyl==6] <- "blue"
x$color[x$cyl==8] <- "darkgreen"

dotchart(x$mpg,
labels = row.names(x),
cex=.7,
groups = x$cyl,
gcolor = "black",
color = x$color,
pch=19,
main = "Gas Mileage for Car Models\ngrouped by cylinder",
xlab = "Miles Per Gallon")
```

