---
title: "Chapter 6: Data Cleaning, Loops, apply functions"
author: "M Affouf"
date: "1/8/2018"
output:output: html_document
---

#part1


## Data

* We will be using multiple data sets in this lecture:
    * Salary, Monument, Circulator, and Restaurant from OpenBaltimore: 
    https://data.baltimorecity.gov/browse?limitTo=datasets
    * Gap Minder - very interesting way of viewing longitudinal data
        * Data is here - http://www.gapminder.org/data/
    * [http://spreadsheets.google.com/pub?key=rMsQHawTObBb6_U2ESjKXYw&output=xls](http://spreadsheets.google.com/pub?key=rMsQHawTObBb6_U2ESjKXYw&output=xls)

## Data Cleaning

In general, data cleaning is a process of investigating your data for inaccuracies, or recoding it in a way that makes it more manageable.

MOST IMPORTANT RULE - LOOK AT YOUR DATA!

Again - `table`, `summarize`, `is.na`, `any`, `all` are useful.


## Dealing with Missing Data

## Missing data types
One of the most important aspects of data cleaning is missing values.  

Types of "missing" data:

* `NA` - general missing data
* `NaN` - stands for "**N**ot **a** **N**umber", happens when you do 0/0.
* `Inf` and `-Inf` - Infinity, happens when you take a positive number (or negative number) by 0.

## Finding Missing data

Each missing data type has a function that returns `TRUE` if the data is missing:

* `NA` - `is.na`
* `NaN` - `is.nan`
* `Inf` and `-Inf` - `is.infinite`
* `is.finite` returns `FALSE` for all missing data and `TRUE` for non-missing
* `complete.cases` on a `data.frame`/`matrix` returns `TRUE` if all values in that row of the object are not missing.

## Missing Data with Logicals

One important aspect (esp with subsetting) is that logical operations return `NA` for `NA` values.  Think about it, the data could be `> 2` or not we don't know, 
so `R` says there is no `TRUE` or `FALSE`, so that is missing:
```{r}
x = c(0, NA, 2, 3, 4)
x > 2
```

## Missing Data with Logicals

What to do?  What if we want if `x > 2` and `x` isn't `NA`?  
Don't do `x != NA`, do `x > 2` and `x` is NOT `NA`:

```{r}
x != NA
x > 2 & !is.na(x)
```

## Missing Data with Logicals

What about seeing if a value is equal to multiple values?  You can do `(x == 1 | x == 2) & !is.na(x)`, but that is not efficient.  Introduce the `%in%` operator:

```{r}
(x == 0 | x == 2) # has NA
(x == 0 | x == 2) & !is.na(x) # No NA
x %in% c(0, 2) # NEVER has NA and returns logical
```

## Missing Data with Operations

Similarly with logicals, operations/arithmetic with `NA` will result in `NA`s:

```{r}
x + 2
x * 2
```
 
## Tables and Tabulations

## Creating One-way Tables {.smaller}

Here we will use `table` to make tabulations of the data.   Look at `?table` to see options for missing data.

```{r table}
table(x)
table(x, useNA = "ifany")
```


## Creating One-way Tables {.smaller}

You can set `useNA = "always"` to have it always have a column for `NA`
```{r onetab}
table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
        useNA = "always")
```

## Creating Two-way Tables {.smaller}

A two-way table.  If you pass in 2 vectors, `table` creates a 2-dimensional table.

```{r}
tab <- table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
             c(0, 1, 2, 3, 2, 3, 3, 4, 4, 3), 
              useNA = "always")
```

## Finding Row or Column Totals

`margin.table` finds the marginal sums of the table.  `margin` is 1 for rows, 2 for columns in general in `R`.  Here is the column sums of the table:

```{r margin}
margin.table(tab, 2)
```

## Proportion Tables {.smaller}

`prop.table` finds the marginal proportions of the table. Think of it dividing the table by it's respective marginal totals.  If `margin` not set, divides by overall total.

```{r table2}
prop.table(tab)
prop.table(tab,1)
```

## Download Salary FY2014 Data

From https://data.baltimorecity.gov/City-Government/Baltimore-City-Employee-Salaries-FY2014/2j28-xzd7

Read the CSV into R `Sal`:

```{r readSal}
Sal = read.csv("Baltimore_City_Employee_Salaries_FY2014.csv",
               as.is = TRUE)
```

## Checking for logical conditions {.smaller}
* `any()` - checks if there are any `TRUE`s
* `all()` - checks if ALL are true

```{r isna}
head(Sal,2)
any(is.na(Sal$Name)) # are there any NAs?
```

# Recoding Variables

## Example of Recoding: base R

For example, let's say gender was coded as Male, M, m, Female, F, f.  Using Excel to find all of these would be a matter of filtering and changing all by hand or using if statements.

In R, you can simply do something like:

```{r, eval = FALSE}
# data$gender[data$gender %in% 
# c("Male", "M", "m")] <- "Male"
```

## Example of Recoding with `recode`: `car` package

You can also `recode` a vector:

```{r}
library(car, quietly = TRUE)
x = rep(c("Male", "M", "m", "f", "Female", "female" ), 
        each = 3)
car::recode(x, "c('m', 'M', 'male') = 'Male';
            c('f', 'F', 'female') = 'Female';")
```

## Example of Recoding with `revalue`: `plyr`

You can also `revalue` a vector with the `revalue` command

```{r}
library(plyr)
plyr::revalue(x, c("M" = "Male", "m" = "Male",
                   "f" = "Female", "female" = "Female"))
```

## Example of Cleaning: more complicated

Sometimes though, it's not so simple.  That's where functions that find patterns come in very useful.

```{r gender, echo=FALSE}
set.seed(4) # random sample below - make sure same every time
gender <- sample(c("Male", "mAle", "MaLe", "M", "MALE", "Ma", "FeMAle", "F", "Woman", "Man", "Fm", "FEMALE"), 1000, replace = TRUE)
```

```{r gentab}
table(gender)
```

# String functions

## Pasting strings with `paste` and `paste0`

Paste can be very useful for joining vectors together:

```{r Paste}
paste("Visit", 1:5, sep = "_")
paste("Visit", 1:5, sep = "_", collapse = " ")
paste("To", "is going be the ", "we go to the store!", sep = "day ")
# and paste0 can be even simpler see ?paste0 
paste0("Visit",1:5)
```

## Paste Depicting How Collapse Works

```{r Paste2}
paste(1:5)
paste(1:5, collapse = " ")
```

## Useful String Functions

Useful String functions

* `toupper()`, `tolower()` - uppercase or lowercase your data:
* `str_trim()` (in the `stringr` package) or `trimws` in base 
    - will trim whitespace
* `nchar` - get the number of characters in a string
* `paste()` - paste strings together with a space 
* `paste0` - paste strings together with no space as default


## The `stringr` package

Like `dplyr`, the `stringr` package:

* Makes some things more intuitive
* Is different than base R
* Is used on forums for answers
* Has a standard format for most functions
    * the first argument is a string like first argument is a `data.frame` in `dplyr`



## Splitting/Find/Replace and Regular Expressions

* R can do much more than find exact matches for a whole string
* Like Perl and other languages, it can use regular expressions.
* What are regular expressions?
    * Ways to search for specific strings 
    * Can be very complicated or simple
    * Highly Useful - think "Find" on steroids


## A bit on Regular Expressions

* http://www.regular-expressions.info/reference.html
* They can use to match a large number of strings in one statement
* `.` matches any single character
* `*` means repeat as many (even if 0) more times the last character
* `?` makes the last thing optional
* `^` matches start of vector `^a` - starts with "a"
* `$` matches end of vector `b$` - ends with "b"

# Splitting Strings

## Substringing

Very similar:

Base R

* `substr(x, start, stop)` - substrings from position start to position stop
* `strsplit(x, split)` - splits strings up - returns list!

`stringr`

* `str_sub(x, start, end)` - substrings from position start to position end
* `str_split(string, pattern)` - splits strings up - returns list!


## Splitting String: base R

In base R, `strsplit` splits a vector on a string into a list
```{r strsplit}
x <- c("I really", "like writing", "R code programs")
y <- strsplit(x, split = " ") # returns a list
y
```

## Splitting String: `stringr`

`stringr::str_split` do the same thing:

```{r str_split}
library(stringr)
y2 <- str_split(x, " ") # returns a list
y2
```

## Using a fixed expression

One example case is when you want to split on a period "`.`".  In regular expressions `.` means **ANY** character, so

```{r}
str_split("I.like.strings", ".")
str_split("I.like.strings", fixed("."))
```

## Let's extract from `y`

```{r stsplit2}
suppressPackageStartupMessages(library(dplyr)) # must be loaded AFTER plyr
y[[2]]
sapply(y, dplyr::first) # on the fly
sapply(y, nth, 2) # on the fly
sapply(y, last) # on the fly
```


## 'Find' functions: base R

`grep`: `grep`, `grepl`, `regexpr` and `gregexpr` search for matches to argument pattern within each element of a character vector: they differ in the format of and amount of detail in the results. 

`grep(pattern, x, fixed=FALSE)`, where:

* pattern = character string containing a regular expression to be matched in the given character vector.

* x = a character vector where matches are sought, or an object which can be coerced by as.character to a character vector.

* If fixed=TRUE, it will do exact matching for the phrase anywhere in the vector (regular find)


## 'Find' functions: `stringr`

`str_detect`, `str_subset`, `str_replace`, and `str_replace_all` search for matches to argument pattern within each element of a character vector: they differ in the format of and amount of detail in the results. 

* `str_detect` - returns `TRUE` if `pattern` is found
* `str_subset` - returns only the strings which pattern were detected
    * convenient wrapper around `x[str_detect(x, pattern)]`
* `str_extract` - returns only strings which pattern were detected, but ONLY the pattern
* `str_replace` - replaces `pattern` with `replacement` the first time
* `str_replace_all` - replaces `pattern` with `replacement` as many times matched

## 'Find' functions: stringr compared to base R

Base R does not use these functions.  Here is a "translator" of the `stringr` function to base R functions

* `str_detect` - similar to `grepl` (return logical)
* `grep(value = FALSE)` is similar to `which(str_detect())`
* `str_subset` - similar to `grep(value = TRUE)` - return value of matched
* `str_replace` - similar to `sub` - replace one time 
* `str_replace_all` - similar to `gsub` - replace many times

## Let's look at modifier for `stringr`

`?modifiers`

* `fixed` - match everything exactly
* `regexp` - default - uses **reg**ular **exp**ressions
* `ignore_case` is an option to not have to use `tolower`


## Important Comparisons

Base R:

* Argument order is `(pattern, x)`
* Uses option `(fixed = TRUE)`

`stringr`

* Argument order is `(string, pattern)` aka `(x, pattern)`
* Uses function `fixed(pattern)`



## 'Find' functions: Finding Indices

These are the indices where the pattern match occurs:

```{r RawlMatch}
grep("Rawlings",Sal$Name)
which(grepl("Rawlings", Sal$Name))
which(str_detect(Sal$Name, "Rawlings"))
```

## 'Find' functions: Finding Logicals

These are the indices where the pattern match occurs:

```{r RawlMatch_log}
head(grepl("Rawlings",Sal$Name))
head(str_detect(Sal$Name, "Rawlings"))
```


## 'Find' functions: finding values, base R {.smaller}

```{r grepl}
grep("Rawlings",Sal$Name,value=TRUE)
Sal[grep("Rawlings",Sal$Name),]
```

## 'Find' functions: finding values, `stringr` and `dplyr` {.smaller}

```{r ggrep}
str_subset(Sal$Name, "Rawlings")
Sal %>% filter(str_detect(Name, "Rawlings"))
```

## Showing differnce in `str_extract`

`str_extract` extracts just the matched string

```{r ggrep2}
ss = str_extract(Sal$Name, "Rawling")
head(ss)
ss[ !is.na(ss)]
```

## Showing differnce in `str_extract` and  `str_extract_all`

`str_extract_all` extracts all the matched strings

```{r }
head(str_extract(Sal$AgencyID, "\\d"))
head(str_extract_all(Sal$AgencyID, "\\d"), 2)
```



## Using Regular Expressions
* Look for any name that starts with:
    * Payne at the beginning, 
    * Leonard and then an S
    * Spence then capital C

```{r grepstar}
head(grep("^Payne.*", x = Sal$Name, value = TRUE), 3)
```

```{r grepstar2}
head(grep("Leonard.?S", x = Sal$Name, value = TRUE))
head(grep("Spence.*C.*", x = Sal$Name, value = TRUE))
```



## Using Regular Expressions: `stringr`
```{r grepstar_stringr}
head(str_subset( Sal$Name, "^Payne.*"), 3)
```

```{r grepstar2_stringr}
head(str_subset( Sal$Name, "Leonard.?S"))
head(str_subset( Sal$Name, "Spence.*C.*"))
```

## Replace

Let's say we wanted to sort the data set by Annual Salary:

```{r classSal}
class(Sal$AnnualSalary)
```

```{r orderstring}
sort(c("1", "2", "10")) #  not sort correctly (order simply ranks the data)
order(c("1", "2", "10"))
```

## Replace

So we must change the annual pay into a numeric:
```{r destringSal}
head(Sal$AnnualSalary, 4)
head(as.numeric(Sal$AnnualSalary), 4)
```

R didn't like the `$` so it thought turned them all to `NA`.

`sub()` and `gsub()` can do the replacing part in base R.

## Replacing and subbing {.smaller}

Now we can replace the `$` with nothing (used `fixed=TRUE` because `$` means ending):

```{r orderSal}
Sal$AnnualSalary <- as.numeric(gsub(pattern = "$", replacement="", 
                              Sal$AnnualSalary, fixed=TRUE))
Sal <- Sal[order(Sal$AnnualSalary, decreasing=TRUE), ] 
Sal[1:5, c("Name", "AnnualSalary", "JobTitle")]
```

## Replacing and subbing: `stringr` {.smaller}

We can do the same thing (with 2 piping operations!) in dplyr
```{r orderSal_stringr}
dplyr_sal = Sal
dplyr_sal = dplyr_sal %>% mutate( 
  AnnualSalary = AnnualSalary %>%
    str_replace(
      fixed("$"), 
      "") %>%
    as.numeric) %>%
  arrange(desc(AnnualSalary))
check_Sal = Sal
rownames(check_Sal) = NULL
all.equal(check_Sal, dplyr_sal)
```


#Part 2

##Agenda

- A common data cleaning task
- If-else statements
- For/while loops to iterate over data
- `apply()`, `lapply()`, `sapply()`, `tapply()`
- `with()` to specify scope

##A common problem

- One of the most common problems you'll encounter when importing manually-entered data is inconsistent data types within columns

- For a simple example, let's look at `TVhours` column in a messy version of the survey data from Lecture 2

```{r}
survey.messy <- read.csv("survey_messy.csv", header=TRUE)
survey.messy$TVhours
```

##What's happening?
```{r}
str(survey.messy)
```
- Several of the entries have non-numeric values in them (they contain strings)

- As a result, `TVhours` is being imported as `factor`

##Attempt at a fix
- What if we just try to cast it back to numeric?

```{r}
tv.hours.messy <- survey.messy$TVhours
tv.hours.messy
as.numeric(tv.hours.messy)
```

##That didn't work...
```{r, eval = FALSE}
tv.hours.messy
as.numeric(tv.hours.messy)
```
```{r, echo = FALSE}
# Show this output, but not this code
head(tv.hours.messy, 40)
head(as.numeric(tv.hours.messy), 40)
```



- This just converted all the values into the integer-coded levels of the factor

- Not what we wanted!

##Something that does work

- Consider the following simple example

```{r}
num.vec <- c(3.1, 2.5)
as.factor(num.vec)
as.numeric(as.factor(num.vec))
as.numeric(as.character(as.factor(num.vec)))
```

> If we take a number that's being coded as a factor and *first* turn it into a `character` string, *then* converting the string to a numeric gets back the number

##Back to the corrupted TVhours column

```{r}
as.character(tv.hours.messy)
as.numeric(as.character(tv.hours.messy))
typeof(as.numeric(as.character(tv.hours.messy)))  # Success!! (Almost...)
```

##A small improvement

- All the corrupted cells now appear as `NA`, which is R's missing indicator

- We can do a little better by cleaning up the vector once we get it to `character` form

```{r}
tv.hours.strings <- as.character(tv.hours.messy)
tv.hours.strings
```

##Deleting non-numeric (or .) characters 
```{r}
tv.hours.strings
# Use gsub() to replace everything except digits and '.' with a blank ""
gsub("[^0-9.]", "", tv.hours.strings) 
```

##The final product
```{r}
tv.hours.messy[1:30]
tv.hours.clean <- as.numeric(gsub("[^0-9.]", "", tv.hours.strings))
tv.hours.clean
```
- As a last step, we should go through and figure out if any of the `NA` values should really be `0`.  
  - This step is not shown here.

##Rebuilding our data
```{r}
survey <- transform(survey.messy, TVhours = tv.hours.clean)
str(survey)
```

- **Success!**

##A different approach

- We can also handle this problem by setting `stringsAsFactors = FALSE` when importing our data.

```{r}
survey.messy <- read.csv("survey_messy.csv", header=TRUE, stringsAsFactors=FALSE)
str(survey.messy)
```

- Now everything is a `character` instead of a `factor`

##One-line cleanup 

- Let's clean up the `TVhours` column and cast it to numeric all in one command

```{r}
survey <- transform(survey.messy, 
                    TVhours = as.numeric(gsub("[^0-9.]", "", TVhours)))
str(survey)
```

##What about all those other `character` variables?

```{r}
table(survey[["Program"]])
table(as.factor(survey[["Program"]]))
```

- Having factors coded as characters may be OK for many parts of our analysis

##To be safe, let's fix things
```{r}
# Figure out which columns are coded as characters
chr.indexes <- sapply(survey, FUN = is.character)
chr.indexes
# Re-code all of the character columns to factors
survey[chr.indexes] <- lapply(survey[chr.indexes], FUN = as.factor)
```

##Here's the outcome
```{r}
str(survey)
```

- **Success!**

##Another common problem

- When data is entered manually, misspellings and case changes are very common

- E.g., a column showing life support mechanism may look like,

```{r}
life.support <- as.factor(c("dialysis", "Ventilation", "Dialysis", "dialysis", "none", "None", "nnone", "dyalysis", "dialysis", "ventilation", "none"))
summary(life.support)
```
###
```{r}
summary(life.support)
```

- This factor has 8 levels even though it should have 3 (dialysis, ventilation, none)

- We can fix many of the typos by running spellcheck in Excel before importing data, or by changing the values on a case-by-case basis later

- There's a faster way to fix just the capitalization issue 


##What are all these [l/s/t/]apply() functions?

- These are all efficient ways of applying a function to margins of an array or elements of a list

- Before we talk about the details of `apply()` and its relatives, we should first understand loops

- **loops** are ways of iterating over data

- The `apply()` functions can be thought of as good *alternatives* to loops 

##For loops: a pair of examples
```{r}
for(i in 1:4) {
  print(i)
}

phrase <- "Good Night, "
for(word in c("and", "Good", "Luck")) {
  phrase <- paste(phrase, word)
  print(phrase)
}
```

##For loops: syntax

> A **for loop** executes a chunk of code for every value of an **index variable** in an **index set**

- The basic syntax takes the form

```{r, eval=FALSE}
# for(index.variable in index.set) {
#   code to be repeated at every value of index.variable
# }
```

- The index set is often a vector of integers, but can be more general

##Example
```{r}
index.set <- list(name="Michael", weight=185, is.male=TRUE) # a list
for(i in index.set) {
  print(c(i, typeof(i)))
}
```

##Example: Calculate sum of each column
```{r}
fake.data <- matrix(rnorm(500), ncol=5) # create fake 100 x 5 data set
head(fake.data,2) # print first two rows

col.sums <- numeric(ncol(fake.data)) # variable to store running column sums
for(i in 1:nrow(fake.data)) {
  col.sums <- col.sums + fake.data[i,] # add ith observation to the sum
}
col.sums

colSums(fake.data) # A better approach (see also colMeans())
```

##while loops

- **while loops** repeat a chunk of code while the specified condition remains true

```{r, eval=FALSE}
day <- 1
num.days <- 365
while(day <= num.days) {
  day <- day + 1
}
```

- We won't really be using while loops in this class

- Just be aware that they exist, and that they may become useful to you at some point in your analytics career

##The various apply() functions

Command | Description
--------|------------
`apply(X, MARGIN, FUN)` | Obtain a vector/array/list by applying `FUN` along the specified `MARGIN` of an array or matrix `X`
`lapply(X, FUN)` | Obtain a list by applying `FUN` to the elements of a list `X`
`sapply(X, FUN)` | Simplified version of `lapply`.  Returns a vector/array instead of list.
`tapply(X, INDEX, FUN)` | Obtain a table by applying `FUN` to each combination of the factors given in `INDEX`  

- These functions are (good!) alternatives to loops

- They are typically *more efficient* than loops (often run considerably faster on large data sets)

- Take practice to get used to, but make analysis easier to debug and less prone to error when used effectively

- You can always type `example(function)` to get code examples (E.g., `example(apply)`)

##Example: apply()
```{r}
colMeans(fake.data)
apply(fake.data, MARGIN=2, FUN=mean) # MARGIN = 1 for rows, 2 for columns
# Function that calculates proportion of vector indexes that are > 0
propPositive <- function(x) mean(x > 0)
apply(fake.data, MARGIN=2, FUN=propPositive) 
```

##Example: lapply(), sapply()
```{r}
lapply(survey, is.factor) # Returns a list
sapply(survey, FUN = is.factor) # Returns a vector with named elements
```

##Example: apply(), lapply(), sapply()
```{r}
apply(cars, 2, FUN=mean) # Data frames are arrays
lapply(cars, FUN=mean) # Data frames are also lists
sapply(cars, FUN=mean) # sapply() is just simplified lapply()
```


##Example: tapply()

- Think of `tapply()` as a generalized form of the `table()` function

```{r}
library(MASS)
# Get a count table, data broken down by Origin and DriveTrain
table(Cars93$Origin, Cars93$DriveTrain)

# Calculate average MPG.City, broken down by Origin and Drivetrain
tapply(Cars93$MPG.city, INDEX = Cars93[c("Origin", "DriveTrain")], FUN=mean)
```

##Example: tapply()

- Let's get the average horsepower by car `Origin` and `Type`

```{r}
tapply(Cars93[["Horsepower"]], INDEX = Cars93[c("Origin", "Type")], FUN=mean)
```

- What's that `NA` doing there?

```{r}
any(Cars93$Origin == "non-USA" & Cars93$Type == "Large")
```

- None of the non-USA manufacturers produced Large cars!

##Example: using tapply() to mimic table()

- Here's how one can use `tapply()` to produce the same output as the `table()` function

```{r}
library(MASS)
# Get a count table, data broken down by Origin and DriveTrain
table(Cars93$Origin, Cars93$DriveTrain)

# This one may take a moment to figure out...
tapply(rep(1, nrow(Cars93)), INDEX = Cars93[c("Origin", "DriveTrain")], FUN=sum) 
```

##with()

- Thus far we've repeatedly typed out the data frame name when referencing its columns

- This is because the data variables don't exist in our working environment

- Using **with**`(data, expr)` lets us specify that the code in `expr` should be evaluated in an environment that contains the elements of `data` as variables

```{r}
with(Cars93, table(Origin, Type))
```

##Example: with()

```{r}
any(Cars93$Origin == "non-USA" & Cars93$Type == "Large")
with(Cars93, any(Origin == "non-USA" & Type == "Large")) # Same effect!
```

```{r}
with(Cars93, tapply(Horsepower, INDEX = list(Origin, Type), FUN=mean))
```

- Using with() makes code simpler, easier to read, and easier to debug


<!-- --- -->
<!-- ```{r} -->
<!--  library(knitr) -->
<!--  purl("chap6s_m.Rmd", documentation = 0) -->
<!-- ``` -->