---
title: "Rmarkdown_practice"
author: "Odlin"
date: "January 23, 2018"
output: html_document
---

#Data
##Data
###Data

The` atmos` data set resides in the `nasaweather` package of the R programming language

Some of the variables in the `atmos` s data set are:

* **temp** - The mean monthly air temperature near the surface of the Earth (measured in degrees kelvin (K))

* **pressure** - The mean monthly air pressure at the surface of the Earth (measured in millibars (mb))

* **ozone** - The mean monthly abundance of atmospheric ozone (measured in Dobson units (DU))

$$fahrenheit=celsius \times \frac{9}{5}+32$$
```{r}
library(nasaweather)
library(dplyr)
library(ggvis)

```


##Prepare Data
```{r}
# The nasaweather and dplyr packages are available in the workspace

# Set the year variable to 1995
year <- 1995

means <- atmos %>%
  filter(year == year) %>%
  group_by(long, lat) %>%
  summarize(temp = mean(temp, na.rm = TRUE),
         pressure = mean(pressure, na.rm = TRUE),
         ozone = mean(ozone, na.rm = TRUE),
         cloudlow = mean(cloudlow, na.rm = TRUE),
         cloudmid = mean(cloudmid, na.rm = TRUE),
         cloudhigh = mean(cloudhigh, na.rm = TRUE)) %>%
  ungroup()

# Inspect the means variable

means
```
```{r}
summary(cars$dist)
```
```{r fi.height=3,fig.width=4}
plot(cars$dist,cars$speed)
```

