---
title: "Time Series & Financial Data"
author: "M. Affouf"
date: "4/10/18"
output: html_document
---
These notes are based on " R for everyone: Advanced Analytics Book by Lander"

### Autoregressive Moving Average
The autoregressive (AR), moving average (MA), or both (ARMA). These models are easy to do in R. The formula for an (ARMA(p,q)) is
$$ X_t -a_1X_{t-1}-\cdots -a_p X_{t-p}=Z_t +t_1Z_{t-1}+\cdots + t_qZ_{t-q}$$
where
$$ Z_t = WN(0,\sigma^2)$$
is white noise.

Make use of the World Bank API to download gross domestic product (GDP) for a number of countries from 1960 to 2017

```{r}
#load the world Bank API package
require(WDI)
# pull the data

gdp<- WDI(country = c("US","CA","GB", "DE", "CN","JP","SG"),indicator = c("NY.GDP.PCAP.CD", "NY.GDP.MKTP.CD"), start=1960, end=2017)
#end=2011)
#give it good names
names(gdp) <- c("iso2c","Country","Year","PerCapGDP","GDP")

#Inspect the data
head(gdp)

library(ggplot2)
library(scales)
#per capita GDP
ggplot(gdp, aes(Year, PerCapGDP, color=Country, linetype=Country))+
  geom_line()+scale_y_continuous(label=dollar)

library(useful)
#absolute GDP
ggplot(gdp, aes(Year, PerCapGDP, color=Country, linetype=Country))+
  geom_line()+scale_y_continuous(label=multiple_format(extra=dollar,multiple="M"))

```

Extract the data for the United states
```{r}
# get US data
us <- gdp$PerCapGDP[gdp$Country == "United States"]
us <- ts(us, start=min(gdp$Year), end = max(gdp$Year))
us

plot(us, ylab = "Per Capita GDP", xlab = "Year")
```

### Getting data
Download the `AT&T` ticker data using the `quantmod` package.


```{r}
library("quantmod")  #Load the quantmod Library
#getSymbols("YHOO",src="google") # from google finance
getSymbols("GOOG",src="yahoo")  # from yahoo finance

getSymbols("AMZN",src="yahoo")  # from yahoo finance
# Specify lookup parameters, and save for future sessions.

getSymbols("AAPL",src="yahoo")

barChart(AAPL) 

# Add multi-coloring and change background to white
candleChart(AAPL,multi.col=TRUE,theme="white")  
```



```{r}
library("quantmod")  #Load the quantmod Library
stockData <- new.env() #Make a new environment for quantmod to store data in

startDate = as.Date("2008-01-13") #Specify period of time we are interested in
#endDate = as.Date("2010-01-12")
endDate = as.Date("2017-10-30")

tickers <- c("COST","ATT") #Define the tickers we are interested in

#Download the stock history (for all tickers)
getSymbols(tickers,env = stockData, src = "yahoo", from = startDate, to = endDate)

#Use head to show first six rows of matrix
head(stockData$COST)

#Lets look at the just the closing prices
Cl(stockData$COST[1:15])

#Lets plot the data
chartSeries(stockData$COST)

#Lets add some bollinger bands to the plot (with period 50 & width 2 standard deviations)

#?addBBands #Make R display the help documentation so we know what variables to pass to the function
addBBands(n=50, sd=2)

#Lets get the technical indicator values saved into a variable
#Note must give it a single time series (I gave it the close price in this example)
indicatorValuesBBands <- BBands(Cl(stockData$COST),n=50, sd=2)

#Lets examine only a 1 month period of data
COSTSubset<-  window(stockData$COST, start = as.Date("2017-10-01"), 
# start = as.Date("2010-02-15"),                    
                     end = as.Date("2017-10-30"))
#as.Date("2010-03-15"))

COSTSubset #Lets see the data

#Lets extract a 1 month period of data for ATT but starting midway through the arm data
ATTSubset<-  window(stockData$ATT, start = as.Date("2010-02-25"), end = as.Date("2010-03-25"))
ATTSubset #Lets see the data

#Now we want to get the intersection of the two subsets of data
#this will gives us all the sets of data where the dates match
#Its important to match the date series to stop spurious analysis of non-synchronised data
#All=FALSE specifies the intersection as in don't include all dates in the merge
COSTATTIntersection <- merge(COSTSubset, ATTSubset, all = FALSE)
#subset(COSTATTIntersection,select = c("COST.Open","ATT.Open")) #Select the open columns and display
```